---
title: "GEOG 365 Exam 1"
date: "September 12, 2019"
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lipsum}
- \pagestyle{fancy}
- \fancyhead[CO]{GEOG 365 Introduction to GIS Programming}
output: pdf_document
---

```{r include = F}
library(knitr)

# Change the default arguments
opts_chunk$set(results = F)
```

# Instructions

Class: Introduction to GIS Programming

Course Code: GEOG 365 

Time: 12:05 PM - 1:20 PM

Total Points: 10

Make sure you follow these instructions:

- Write down your name and PSU username (e.g. xys123) in the section below.
- Prior to each question, the number of points is shown, e.g. 1' for 1 point.
- Turn in the entire exam when you are finished.
- Handwriting should be clear and legible.
- Each question contains subquestions. Ensure you answer all questions.
- If you run out of space, you can write on the reverse of the page and state the question number. 

# Personal Information

First Name: \_\_\_\_\_\_\_\_\_\_

Last Name: \_\_\_\_\_\_\_\_\_\_

PSU username: \_\_\_\_\_\_\_\_\_\_

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

| **DO NOT start the exam until you are told to do so!** |
|:-----------:|

\newpage

# Questions

1. A matrix is created as follows:

```{r}
vec <- c(1, 5, 9, 2, 6, 10, 3, 7, 11, 4, 8, 12)
mat <- matrix(vec, ncol = 4)
```

1.1 (0.4') Write down the content of the matrix *mat* in a 2-dimensional table form (hint: similar to what *print(mat)* would generate).

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

1.2 (0.3') What is the order style in storage of the matrix?

&nbsp;

&nbsp;

1.3 (0.3') After the creation of the matrix, consider the command *length(mat)*. What response would you expect from running the command in R?

&nbsp;

&nbsp;

2. The following questions test your knowledge on date-time object handling in R.

2.1 (0.5') Consider the following code snippet.

```{r}
z1 <- c(43200, 86400, 108000)
dt1 <- as.POSIXct(z1, origin = '2019-01-20', tz = 'UTC')
```

Will the above code run successfully? If so, write down the content of the vector *dt1* in the form *YYYY-mm-dd HH:MM:SS TimeZone*; if not, explain.

&nbsp;

&nbsp;

&nbsp;

&nbsp;

2.2 (0.5') Consider the following code snippet.

```{r}
z2 <- c('02/03/2019', '03/03/2019')
dt2 <- strptime(z2, format = "%d/%m/%Y", tz = 'UTC')
print(dt2[2] - dt2[1])
```

Write down the result of the *print* function.

\newpage

3. Consider the following scenario: you just loaded a data frame *df* into your R environment; all that you know about this data frame is that it contains daily temperature measurements for Pennsylvania.

3.1 (0.5') Mention at least 3 functions that you could use to examine the different attribtues of the data frame. Explain what attribute each function will examine. No points will be given if you do not explain.

&nbsp;

&nbsp;

&nbsp;

&nbsp;

3.2 (0.5') The data frame contains the original measurements. Describe what you would check before applying statistical modelling and analysis techniques.

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

4. Consider the following code snippet:

```{r}
vec1 <- seq_len(4)
vec2 <- 4:6
vec.new <- c()

for (i in 1:length(vec1)) {
	for (j in 1:length(vec2)) {
		vec.new <- c(vec.new, vec1[i] * vec2[j])
	}
}
```

4.1 (1') Write down the content of the vector *vec.new*.

&nbsp;

&nbsp;

&nbsp;

4.2 (1') However, the programmer is not very careful when typing the code. The line

*for (i in 1:length(vec1)) {*

becomes

*for (i in length(vec1)) {*

Is the content of *vec.new* going to change? Explain why. If the content is going to change, also write down the content after the change.

\newpage

5. Read the following remarks on the recycling mechanism in the R language.

Recycling is when an object is automatically extended in certain operations to match the length of another, longer object. For example, the addition operation generates the following result:

```{r results = T}
c(1, 2, 3) + c(1, 2, 3, 4, 5, 6)
```

Because of the recycling, the operation that actually happened is:

```{r}
c(1, 2, 3, 1, 2, 3) + c(1, 2, 3, 4, 5, 6)
```

In cases where the longer object is not a multiple of the shorter one, a warning message is presented:

```{r results = T}
c(1, 2, 3) + c(1, 2, 3, 4)
```

5.1 (1') Write down the result and warning messages (if any) of the command *1:4 \* 2:3*.

&nbsp;

&nbsp;

5.2 (1') Write down the result and warning messages (if any) of the command

*seq(from = 2, length.out = 4, by = 2) / c(2, 0, 1)*

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

5.3 (1') Write down the result and warning messages (if any) of the command

*matrix(1:5, nrow = 2, ncol = 5)*

\newpage

6. Consider the following function definition.

```{r}
func <- function(vec, a = 1, b = 2, n = 1) {
	
	for (v in vec) 
	if (v == 0)
		cat('The operation on 0 is ignored!\n')
	else {
		result <- a * v ^ n + log(v, base = b)
		cat(result, '\n')
	}
}
```

6.1 (0.3') Will the command *func(c(4, 8, 0, 16))* run successfully? If so, write down the output messages; if not,  explain.

&nbsp;

&nbsp;

&nbsp;

6.2 (0.3') Will the command *func(4, 8, 2, 2)* run successfully? If so, write down the output messages; if not,  explain.

&nbsp;

&nbsp;

&nbsp;

6.3 (0.4') Will the command *func(a = 3)* run successfully? If so, write down the output messages; if not,  explain.

&nbsp;

&nbsp;

&nbsp;

6.4 (1') If you were going to write the function *func*, how would you improve the **coding style**? First, write down the improvements you plan to make on the left. Then, modify the above function code on the right. **Make sure you describe the improvements you plan to make on the left otherwise I would not be able to discover all your improvements purely based on your writing from the right**.

Description of Improvements
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
Your Implementation
